Covariate analysis of ROSMAP reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-03-22 code/Rmd/ROSMAP_REPROCESSED.Rmd
Modified ROSMAP reprocessing to include new picard metrics and sva processing

@@

# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8380366'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)

@@

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples with no cogdx, RIN and PMI scores
METADATA <- METADATA %>%
  ungroup %>%
  filter(Sampleid %in% colnames(COUNT)) %>%
  filter(!is.na(cogdx)) %>%
  filter(!is.na(RINcontinuous)) %>%
  filter(!is.na(pmi)) %>%
  filter(!is.na(Sampleid)) %>%
  filter(!(Sampleid %in% c('380_120503'))) %>%
  as.data.frame()
```

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE", "age_first_ad_dx","age_at_visit_max",
primaryVariable <- "cogdx"
FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ", "PCT_PF_READS", "PCT_PF_READS_ALIGNED",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")

@@

# Calculate the number of surrogate variables
NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                     MODEL1, method = 'be', seed = 123456, B = 30)

# Estimate surrogate variables
SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv

@@

# Store differential expression results
write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Results (cogdx)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

>>>>

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

@@

# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8512191'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)

@@

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples with no cogdx, RIN and PMI scores
METADATA <- METADATA %>%
  ungroup %>%
  filter(Sampleid %in% colnames(COUNT)) %>%
  filter(!is.na(cogdx), !is.na(braaksc), !is.na(ceradsc)) %>%
  filter(!is.na(RINcontinuous)) %>%
  filter(!is.na(pmi)) %>%
  filter(!(Sampleid %in% c('380_120503'))) %>%
  as.data.frame()

METADATA$Diagnosis = 'NONE'
METADATA$Diagnosis[METADATA$cogdx == 1 & METADATA$braaksc <= 3 & METADATA$ceradsc > 2] = 'Control'
METADATA$Diagnosis[METADATA$cogdx == 4 & METADATA$braaksc >= 4 & METADATA$ceradsc <= 2] = 'AD'
```

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE", "age_first_ad_dx","age_at_visit_max",
primaryVariable <- "cogdx"
FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")

@@

# Calculate the number of surrogate variables (be method)
# NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, MODEL1, method = 'be', seed = 123456, B = 30)

# Get residuals from differential expression model
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, VOOM.WEIGHTS$E)

# Compute actual variance of all principal components
tmp = svd(RESIDUAL.SVA.GENE_EXPRESSION)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
  tmp.residual = t(apply(RESIDUAL.SVA.GENE_EXPRESSION, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

# Plot variance components
var.comp = data.frame(component = 1:length(actual.var), 
                      residual = actual.var, 
                      permuted.residual = permuted.var) %>%
  tidyr::gather(data, value, -component)
p = ggplot(var.comp %>% filter(component <= 10), aes(x = component, y = value, color = data)) + geom_point()
p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
p

# Estimate surrogate variables
SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv
SURR.VAR = data.frame(SURR.VAR)

@@

# Store differential expression results
write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Results (cogdx)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
